The Noisy Work of Uncertainty Visualisation Research: A Review

### 2024-04-23 paper.qmd
added small comment in grammar of graphics section
addded some hand drawn graphics

@@

![(Notice: Need to replace with a more relevant data pipeline that includes all the uncertainty stuff and how uncertainty arises at each stage.) The grammar of graphics data analysis pipeline from @Leland2005. the "varset" stages are the mathematical components of the graphic, the mathematical components define the "information" in the graph, while the "graph" stages are the visual elements that determine the "computational" cost of the graph.](grammarofgraphicspipeline.png){#fig-pipeline}

 ![(infintiegraph](infinitegraph.png){#fig-infinitegraph}

@@

#### Decision making and risk aversion
Decision making tasks are often described by authors as a more realistic version how plots are used in practice. While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained. While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots. Let us consider this in more detail. A normal value extraction task involves:

@@

### Just noticeable signal
It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered. The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework. Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.

![Notice: I will fix this image up, it is currently a quick scribble. Difference between the null plot vs the uncertainty plot. A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference) while an uncertainty plot displays the variance of the estimated value and assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero).](hypvs.jpeg){#fig-hypvs}

>>>>

@Tierney2023 generates a visualisation for missing values, however because it fits nicely into the grammar of graphics framework, it is inherrently a descriptive statistic.

@@

![(The grammar of graphics data analysis pipeline from @Leland2005 written in black, with the tidy data equivalent written in blue. The product plot framework and the grammar of probabilistic graphics frameworks are in green. It can be seen that the grammar of graphics provides a set of instructions that create a graphic, however the inferential power of this graphic is not considered and relies on many additional assumptions that are not defined by the grammar.](grammarofgraphicspipeline.jpeg){#fig-pipeline}

 ![This graphic shows three simple graphics made from the same data set that diverge according to the grammar of graphics. Despite being distinct graphics from a stage as low down as the statistic (or data manipulation in the tidy data framework), as our sample size increases to infinity, the grahics converge to the same distribution and become visually indistinguishable. ](infinitegraph.jpeg){#fig-infinitegraph}

@@

#### Decision making and risk aversion

![Visualisation showing the difference ](extractvsdecision.jpeg){#fig-extract}

Decision making tasks are often described by authors as a more realistic version how plots are used in practice. While we do not disagree with this statement, exactly *why* these tasks are different to value extraction tasks is never explained. While the task may be more similar to how plots are actually used, that does not necessarily mean it is a better *experimental* environment for evaluating plots. Let us consider this in more detail. A normal value extraction task involves:

@@

### Just noticeable signal
It could be argued that a well done uncertainty visualisations should have an imperceptible signal unless the signal would be identified with a hypothesis test, almost like a reverse line up protocol, but this idea also has some issues that should be considered. The reject or do not reject concepts in hypothesis testing do not offer a complete image of uncertainty, and exploration of uncertainty visualisation largely stems from a desire to move away from this binary framework.

![The visualisation showing the kind of signal that we would expect an uncertainty plot to supress](signalsupression.jpeg){#fig-sigsupress}

Additionally, setting up an uncertainty visualisation where the participants are expected to notice the signal once the data behind the visualisation passes a hypothesis test implies the signal *is* noticeable to a human at that level. @Patrick2023 compared people ability to recognise patterns in a residual plot to typical statistical tests and found human viewers looking at a plot were less sensitive than the typical residual tests. These experiments utilised the line-up protocol which has been suggested as a method to check if perceived patterns are real or merely the result of chance [@Buja2009; @Wickham2010; @Chowdhury]. This concept bears similarity to the goal of uncertainty visualisation, but it is not quite the same. @fig-hypvs shows the conceptual difference between the line-up protocol and uncertainty visualisation.

![A line-up protocol displays the uncertainty about the null and identifies if the true data plot is identifiable (and therefore significantly difference). This can be considered the graphical equivalent of a standard hypothesis test. Uncertainty is frequently used to descibe the area around the estimate that we think might contain the true value, and we assesses if the null of "no signal" is within this plot (e.g. if the error bars overlap with zero). Uncertainty is technically the distance between the true value and the estimate. ](hypvs.jpeg){#fig-hypvs}
